---
title: "Practical 4 - Tidyverse"
---

::: {.callout-tip}
## Note on Tables
Most tables have been limited to 300px in height and are scrollable, in the interest of a more readable website.
:::

## Question 1
```{r}
#| echo: true
#| message: false
library(tidyverse)
library(nycflights13)
library(knitr)
library(kableExtra)

glimpse(flights)
```

## Question 2
```{r}
#| echo: true
#| message: false
dist_tbl <- 
  flights %>%
  filter(month == 1) %>%
  group_by(carrier) %>%
  summarise(
    mean_dist = mean(distance, na.rm = TRUE),
    sd_dist   = sd(distance, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_dist))

dist_tbl %>%
  kable() %>%
  scroll_box(width = "100%", height = "300px")
```

## Question 3
### Question 3.1
The 'OO' carrier contains only a single observation for Month 1; consequently, the standard deviation is undefined (NA), as a minimum of two data points is required to calculate variance.
```{r}
#| echo: true
#| message: false
check1 <- 
  flights %>%
  filter(carrier == "OO") %>%
  filter(month == 1)

kable(check1)
```

### Question 3.2
For carriers F9, YV, AS, and HA, the flight distances in Month 1 were uniform. Because there is no variance between these observations, the resulting standard deviation is zero.
```{r}
check2 <- 
  flights %>%
  filter(carrier %in% c("F9", "YV", "AS", "HA")) %>%
  filter(month == 1) %>%
  select(distance, month, carrier) %>% 
  arrange(desc(distance))

check2 %>% 
  kable() %>%
  scroll_box(width = "100%", height = "300px")
```

## Question 4
```{r}
#| echo: true
#| message: false
dep_tbl <- 
  flights %>%
  summarise(
    mean_dep = mean(dep_delay, na.rm = TRUE),
    .by = c(carrier, month)
  ) %>%
  pivot_wider(
    names_from = carrier,
    values_from = mean_dep
  ) %>%
  arrange(month)

kable(dep_tbl)
```

## Question 5
```{r}
#| echo: true
#| message: false
fast_flights <- 
  flights %>% 
  mutate(speedy = dep_delay > 0 & arr_delay <= 0) %>% 
  summarise(
    speedy_count = sum(speedy, na.rm = T),
    speedy_prop = speedy_count/length(speedy)*100
  )

kable(fast_flights)
```
## Question 6
### Question 6.1
```{r}
#| echo: true
#| message: false
mult_routes <- flights %>% 
  mutate(route = paste(origin, dest, sep = "-")) %>% 
  group_by(route) %>% 
  summarise(
    num_carriers = n_distinct(carrier)
  ) %>% 
  filter(num_carriers > 1) %>%        
  arrange(desc(num_carriers))

mult_routes %>%
  kable() %>%
  scroll_box(width = "100%", height = "300px")
```

### Question 6.2
```{r}
#| echo: true
#| message: false
details <- 
  flights %>% 
  mutate(route = paste(origin, dest, sep = "-")) %>% 
  filter(route %in% mult_routes$route) %>% 
  left_join(airlines, by = "carrier") %>%
  summarise(
    ave_delay = mean(arr_delay, na.rm = T),
    .by = c(carrier, route, name)
  ) %>% 
  arrange(desc(ave_delay))

details %>%
  kable() %>%
  scroll_box(width = "100%", height = "300px")
```

### Question 6.3
```{r}
#| echo: true
#| message: false
details_routes <- details %>% 
  summarise(
    min_delay = min(ave_delay),
    carrier_min = carrier[which.min(ave_delay)],
    
    max_delay = max(ave_delay),
    carrier_max = carrier[which.max(ave_delay)],
    
    .by = route
  )

details_routes %>% 
  kable() %>%
  scroll_box(width = "100%", height = "300px")
```

### Question 6.4
```{r}
#| echo: true
#| message: false
worst_diff <-
  details_routes %>% 
  summarise(
    delta = max_delay - min_delay,
    .by = route
  ) %>% 
  arrange(desc(delta))

worst_diff %>% 
  kable() %>%
  scroll_box(width = "100%", height = "300px")
```

### Question 6.5
The high average delay for ExpressJet is driven by a single outlier event on the JFK to ATL route. Because the dataset for this specific path consists of only one observation, the average is entirely dictated by that solitary delay, thus leading to it becoming the worst difference
```{r}
#| echo: true
#| message: false
reason <- 
  flights %>% 
  mutate(route = paste(origin, dest, sep = "-")) %>%
  filter(route == "JFK-ATL") %>% 
  summarise(
    ave_delay = mean(arr_delay, na.rm = T),
    .by = carrier
  )  
kable(reason)

reason2 <- 
  flights %>% 
  mutate(route = paste(origin, dest, sep = "-")) %>%
  filter(route == "JFK-ATL", carrier == "EV")
kable(reason2)
```

## Question 7
* **Typographical Errors:** The `Gender` variable contains a misspelling ("femal").
* **Redundant Factor Levels:** `Disease status` contains duplicates ("healthy"/"Healthy").
* **Missing Data:** `Glucose` and `Height` contain `NA` entries.

```{r}
#| echo: true
#| message: false
#| warning: false
data <- structure(list(id = c("id_1", "id_2", "id_3", "id_4", "id_5", 
"id_6", "id_7", "id_8", "id_9", "id_10", "id_11", "id_12", "id_13", 
"id_14", "id_15", "id_16", "id_17", "id_18", "id_19", "id_20", 
"id_21", "id_22", "id_23", "id_24", "id_25", "id_26", "id_27", 
"id_28", "id_29", "id_30", "id_31", "id_32", "id_33", "id_34", 
"id_35", "id_36", "id_37", "id_38", "id_39", "id_40", "id_41", 
"id_42", "id_43", "id_44", "id_45", "id_46", "id_47", "id_48", 
"id_49", "id_50"), age = c(50L, 34L, 70L, 33L, 22L, 61L, 69L, 
73L, 62L, 56L, 71L, 33L, 73L, 44L, 45L, 46L, 24L, 70L, 46L, 76L, 
47L, 76L, 28L, 48L, 54L, 27L, 45L, 26L, 61L, 28L, 38L, 55L, 33L, 
36L, 62L, 58L, 72L, 31L, 34L, 51L, 61L, 64L, 26L, 28L, 60L, 29L, 
42L, 46L, 79L, 72L), gender = c("male", "male", "male", "female", 
"female", "male", "female", "male", "male", "female", "female", 
"male", "male", "female", "male", "male", "male", "male", "female", 
"male", "male", "male", "male", "female", "femal", "male", "female", 
"female", "female", "female", "male", "female", "female", "female", 
"male", "male", "female", "male", "female", "female", "male", 
"female", "female", "male", "male", "female", "male", "male", 
"male", "female"), height = c(174.4, 197.7, 174.1, 194.5, NA, 
180.4, 170.5, 157.4, 196.8, 165.1, 153, 197.4, 186, 157.1, 177.5, 
197.7, 179.3, 170.2, 182.4, NA, 165.4, 161, 168.5, 199.2, 157.7, 
154.6, 157.1, 184.5, 181, 194.6, 183.6, 186.9, 176.1, 183, 191.1, 
189.3, 199, 172, 165.6, 170.5, 150.5, 159.2, 192.1, 161.6, 162, 
153.8, 162.3, 186.6, 192.4, 174.9), weight = c(69.4, 62.3, 55.6, 
69.5, 78.6, 60.8, 72.2, 60.9, 75.1, 67.7, 82.5, 68.7, 67.8, 76.7, 
87, 61.1, 70.6, 63.3, 81.5, 59.2, 93.2, 87.3, 83.4, 80.9, 68.6, 
76.5, 93.7, 79.1, 92, 65.6, 85.4, 63.3, 79.7, 74.1, 63.3, 78.2, 
95.7, 95.1, 63.7, 66.1, 99.3, 81, 96.9, 73.3, 70.3, 83, 57.6, 
78.6, 61.9, 98.1), blood_type = c("O", "A", "O", "O", "B", "AB", 
"O", "O", "O", "AB", "A", "O", "O", "O", "B", "A", "B", "AB", 
"O", "AB", "A", "AB", "O", "B", "A", "A", "B", "AB", "A", "B", 
"B", "A", "O", "O", "O", "B", "O", "A", "A", "B", "A", "O", "AB", 
"A", "A", "O", "O", "B", "A", "O"), disease_status = c("diseased", 
"healthy", "healthy", "healthy", "healthy", "healthy", "diseased", 
"healthy", "diseased", "Healthy", "diseased", "healthy", "diseased", 
"healthy", "diseased", "healthy", "healthy", "healthy", "healthy", 
"healthy", "healthy", "diseased", "healthy", "diseased", "healthy", 
"healthy", "healthy", "healthy", "diseased", "diseased", "healthy", 
"healthy", "healthy", "diseased", "diseased", "diseased", "healthy", 
"diseased", "healthy", "healthy", "healthy", "healthy", "healthy", 
"diseased", "diseased", "diseased", "healthy", "healthy", "diseased", 
"diseased"), cholesterol = c(228, 223, 213, 198, 166, 151, 195, 
199, 189, 196, 221, 156, 185, 230, 234, 174, 185, 236, 235, 180, 
165, 220, 160, 153, 250, 153, 184, 242, 212, 179, 224, 233, 181, 
199, 220, 214, 214, 248, 191, 162, 203, 173, 199, 187, 248, 189, 
173, 212, 164, 247), glucose = c(96, 78, 101, 119, 103, 91, 86, 
NA, 77, 80, 115, 85, 88, 109, NA, 71, 90, 94, 91, 87, 113, 93, 
97, 118, 109, 80, 85, 119, 99, 108, 89, 108, 97, 116, 79, 84, 
75, 81, 119, NA, 106, 109, 75, 82, 84, 75, 76, 120, 119, 77), 
    smoker = c("yes", "yes", "yes", "yes", "no", "yes", "no", 
    "yes", "no", "no", "no", "no", "no", "yes", "no", "yes", 
    "yes", "yes", "yes", "yes", "yes", "yes", "yes", "yes", "no", 
    "no", "yes", "yes", "yes", "no", "no", "yes", "no", "yes", 
    "no", "yes", "no", "yes", "yes", "yes", "no", "no", "yes", 
    "no", "no", "no", "no", "no", "no", "yes"), exercise = c("occasional", 
    "regular", "occasional", "regular", "none", "occasional", 
    "regular", "none", "occasional", "none", "occasional", "none", 
    "none", "regular", "occasional", "none", "regular", "regular", 
    "none", "occasional", "none", "occasional", "occasional", 
    "occasional", "regular", "occasional", "regular", "regular", 
    "regular", "occasional", "occasional", "none", "none", "regular", 
    "occasional", "occasional", "none", "none", "none", "none", 
    "occasional", "regular", "regular", "none", "regular", "occasional", 
    "occasional", "none", "occasional", "regular"), income = c(84820L, 
    81547L, 22588L, 72490L, 74533L, 25338L, 41469L, 57315L, 63629L, 
    88662L, 62615L, 56261L, 58499L, 82232L, 77584L, 77275L, 38468L, 
    54510L, 91326L, 78611L, 31402L, 29586L, 21441L, 58269L, 84173L, 
    88295L, 37940L, 43750L, 69750L, 92356L, 82518L, 91455L, 68866L, 
    51178L, 68275L, 27689L, 35418L, 81318L, 62405L, 86851L, 25654L, 
    47553L, 74474L, 51409L, 22607L, 55360L, 96351L, 21516L, 41927L, 
    55810L), education = c("master", "bachelor", "PhD", "master", 
    "bachelor", "highschool", "PhD", "highschool", "PhD", "PhD", 
    "bachelor", "highschool", "master", "bachelor", "PhD", "PhD", 
    "PhD", "bachelor", "master", "highschool", "PhD", "highschool", 
    "bachelor", "master", "highschool", "highschool", "master", 
    "master", "bachelor", "PhD", "highschool", "PhD", "master", 
    "master", "master", "PhD", "highschool", "master", "master", 
    "highschool", "bachelor", "highschool", "bachelor", "PhD", 
    "bachelor", "highschool", "master", "highschool", "bachelor", 
    "bachelor"), region = c("North", "South", "North", "West", 
    "North", "West", "South", "South", "West", "South", "West", 
    "South", "West", "East", "North", "West", "North", "North", 
    "West", "North", "East", "West", "South", "North", "North", 
    "East", "East", "North", "North", "West", "South", "West", 
    "West", "East", "West", "North", "West", "North", "East", 
    "North", "West", "South", "South", "East", "North", "West", 
    "West", "East", "North", "East"), marital_status = c("divorced", 
    "single", "divorced", "divorced", "divorced", "divorced", 
    "divorced", "married", "divorced", "married", "divorced", 
    "widowed", "married", "single", "widowed", "widowed", "single", 
    "divorced", "widowed", "widowed", "single", "married", "single", 
    "married", "widowed", "married", "single", "single", "widowed", 
    "married", "widowed", "divorced", "single", "married", "single", 
    "widowed", "widowed", "married", "widowed", "divorced", "married", 
    "married", "divorced", "single", "married", "widowed", "divorced", 
    "divorced", "single", "divorced")), row.names = c(NA, -50L
), class = c("tbl_df", "tbl", "data.frame"))

UtilsDataRSV::view_cols(data)
```

